Conflict-Driven Evolution and Urban Development: \\ Evidence from the Holy Roman Empire\footnote{ I thank Lara Andermann, Leonie Bielefeld, Daniel Bursian, Davide Cantoni, Emilio Esguerra, Filip Milojevic, Johannes Vorbach and Matthias Weigand for helpful comments and feedback. }

### 2023-06-27 paper/paper.tex
#32 final commit after submission

\title{Conflict Driven Evolution and Urban Development: \\ Evidence from the Holy Roman Empire\footnote{
I thank Lara Andermann, Leonie Bielefeld, Daniel Bursian, Davide Cantoni, Emilio Esguerra, Filip Milojevic, Johannes Vorbach and Matthias Weigand for helpful comments and feedback.
}
}
\author{Elias Hadj Ammar}
\date{June 27, 2023}

@@

\begin{center}
    {\centering\huge{{Conflict-Driven Evolution and Urban Development:\\
Evidence from the Holy Roman Empire
}}}
\end{center}
\date{\vspace{-5ex}}
\thispagestyle{empty}

@@

In this paper I study the role of this process in the economic development of the Holy Roman Empire. In particular, I estimate the effect of state power, the trait that conflict selects for, on urban development. If the effect is positive, then the conflict driven evolution of states may partially explain the eventual ascendancy of Europe: by selecting for military success, Europe's violent history may have accidentally produced fiscal capacity, efficient legal systems, and prospering middle classes.

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win them with state power, and state power is simply aggregate production in excess of the subsistence level. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote
{
The basic idea of Darwinian selection has previously been borrowed by economists - for example by \cite{galor2002} who argue that humans are subject to selection and accumulate (economically) favourable traits over the course of all history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that at some point the European rich began to out-breed the poor and passed their middle-class values on to their offspring.
}

@@

\cite{diamond1997} and \cite{landes1969, landes2006} have argued for the economic benefits of political fragmentation and state competition. \cite{diamond1997} asserts that geography 
remoteness of China deprived it of the benefits of state competition that European states enjoyed throughout their history. \cite{landes2006} makes a similar argument, pointing out how the totalitarianism and bureaucratism that stifled the spread of innovation in medieval China would not have survived long in a more fragmented state system. Geography and fragmentation also play central roles in \cite{cervellati2022}'s theory of European state history.

@@

I justify this assumption with the fact that a conqueror is, by definition, the winner of an armed conflict. State power is not the only determinant of winning armed conflicts, though: luck plays a central role, as do many other time-varying factors for which I cannot control. On the flip side, winning armed conflicts is also not the only consequence of state power. Importantly, the rulers of more powerful states may be able to make more advantageous matches for their children --- dynastic marriages that eventually result in the expansion of their territory. Exhibit A in this argument is the House of Habsburg: they became a dominant power in Europe at the turn of the 16th century when strategic marriages made them rulers of Burgundy and Castille, but Habsburgs had been the Holy Roman Emperors for decades prior.

@@

While the research question may not seem hugely relevant for the present-day Western world, I want to conclude by talking, briefly, about the Democratic Republic of the Congo (DRC). This African state has a weak central government and large parts of its territory are controlled by independent armed groups that are in many ways like small proto-states. \citep[p. 34]{sierra2020}. Recent research has studied state formation \citep{sierra2020}, taxation and state capacity \citep{balan2022}, and conflict \citep{henn2021} in the setting of the DRC. One could argue that we are witnessing economic history "in real time" there.\footnote
{
This thought is not my own; I got it from Davide Cantoni.
} 
In this sense a better understanding of how the European state system evolved may not only deepen our knowledge of the origins of institutions --- it could also inform our perspective, and those of policy makers, on regions of the world that do not have a stable state system at present.

>>>>

\title{Conflict-Driven Evolution and Urban Development: \\ Evidence from the Holy Roman Empire\footnote{
I thank Lara Andermann, Leonie Bielefeld, Daniel Bursian, Davide Cantoni, Emilio Esguerra, Filip Milojevic, Johannes Vorbach and Matthias Weigand for helpful comments and feedback.
}
}
\author{Elias Hadj Ammar}
\date{June 27, 2023}

@@

\begin{center}
    {\centering\Large{{Conflict-Driven Evolution and Urban Development:\\

Evidence from the Holy Roman Empire
}}}
\end{center}
\date{\vspace{-5ex}}
\thispagestyle{empty}

@@

In this paper I study the role of this process in the economic development of the Holy Roman Empire. In particular, I estimate the effect of state power, the trait that conflict selects for, on urban development. If the effect is positive, then the conflict-driven evolution of states may partially explain the eventual ascendancy of Europe: by selecting for military success, Europe's violent history may have accidentally produced fiscal capacity, efficient legal systems, and prospering middle classes.

I build my paper on the theory of conflict-driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win them with state power, and state power is simply aggregate production in excess of the subsistence level. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote
{
The basic idea of Darwinian selection has previously been borrowed by economists - for example by \cite{galor2002} who argue that humans are subject to selection and accumulate (economically) favourable traits over the course of all history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that at some point the European rich began to out-breed the poor and passed their middle-class values on to their offspring.
}

@@

\cite{diamond1997} and \cite{landes1969, landes2006} have argued for the economic benefits of political fragmentation and state competition. \cite{diamond1997} asserts that the geographic remoteness of China deprived it of the benefits of state competition that European states enjoyed throughout their history. \cite{landes2006} makes a similar argument, pointing out how the totalitarianism and bureaucratism that stifled the spread of innovation in medieval China would not have survived long in a more fragmented state system. Geography and fragmentation also play central roles in \cite{cervellati2022}'s theory of European state history.

@@

I justify this assumption with the fact that a conqueror is, by definition, the winner of an armed conflict. State power is not the only determinant of winning armed conflicts, though: luck plays a central role, as do other time-varying factors for which I cannot control. On the flip side, winning armed conflicts is also not the only consequence of state power. Importantly, the rulers of more powerful states may be able to make more advantageous matches for their children --- dynastic marriages that eventually result in the expansion of their territory. Exhibit A in this argument is the House of Habsburg: they became a dominant power in Europe at the turn of the 16th century when strategic marriages made them rulers of Burgundy and Castille, but Habsburgs had already been the Holy Roman Emperors for decades prior.

@@

While the research question may not seem hugely relevant for the present-day Western world, I want to conclude by talking, briefly, about the Democratic Republic of the Congo (DRC). This African state has a weak central government and large parts of its territory are controlled by independent armed groups that are in many ways like small proto-states \citep[p. 34]{sierra2020}. Recent research has studied state formation \citep{sierra2020}, taxation and state capacity \citep{balan2022}, and conflict \citep{henn2021} in the setting of the DRC. One could argue that we are witnessing economic history "in real time" there.\footnote
{
This thought is not my own; I got it from Davide Cantoni.
} 
In this sense a better understanding of how the European state system evolved may not only deepen our knowledge of the origins of institutions --- it could also inform our perspective, and those of policy makers, on regions of the world that do not have a stable state system at present.
